A Test of the Frequency Independence Assumption of the Power System Parameters used in Geomagnetically Induced Current Calculations

### 2019-02-01 paper.tex

Geomagnetically induced currents (GICs) are electric currents in man-made conducting systems due to electric fields induced near Earth's surface. These fields are a result of time variations in ionospheric currents on timescales on the order of hours or less \citep{Ohtani2000} and the movement of Earth's surface relative to slow-varying current systems in Earth's ionosphere that are near stationary relative to the Sun \citep{Stening2013}. Of particular interest are currents induced in electric power systems as they can lead to system degradation, disruption, and failure \citep{Albertson1993,NERC2012}. Accurate estimation of GICs given direct measurements of the electric field (or estimates based on the geomagnetic field) is important for power system design, retrospective analysis, and for the mitigation of the impacts of space weather on power systems [REFS].

A common assumption made in estimates of GIC using either measured or estimated values of the horizontal geoelectric field at Earth's surface, $\mathbf{E}$, is that the power system is resistive or quasi-DC so that the relationship $GIC(t) = a_oE_x(t) + b_oE_y(t)$ holds \citep{Albertson1981,Lehtinen1985}.  The coefficients $a_o$ and $b_o$ can then be calculated analytically using information about the connectivity of the power lines and values of the conductor and transformer resistances using DC circuit methods \citep[e.g.][]{Boteler2014a,Boteler2014b}. This assumption has been used or stated in many works in the past decade, for example, \citet{Pulkkinen2007,Wik2008,Pulkkinen2010,Ngwira2011,Viljanen2012,Overbye2012}.

@@

The 1-second-cadence GIC data used in this paper span a total of 36 days across the 10 intervals: 2006-04-02 -- 2006-04-10, 2006-07-25 -- 2006-07-29, 2006-08-05 -- 2006-08-08, 2006-08-19 -- 2006-08-21, 2006-11-07 -- 2006-11-12, 2006-11-28 -- 2006-11-29, 2006-12-01, 2006-12-14 -- 2006-12-15, 2007-11-18 -- 2007-11-20, and 2007-11-22 -- 2007-11-22 that have been presented previously in \citet{Watari2009} and \cite{Watari2015}.

@@

Unphysical spikes in the electric field measurements were concluded when the absolute value of a component $i$ of the electric field changed by more than 1 mV/km at time $t_s$, and the values $E_i(t_s-2)$ through $E_i(t_s+4)$ were replaced with a linear interpolation of the values at $E_i(t_s-3)$ and $E_i(t_s + 5)$. In the sample shown in Figure~\ref{figure1}, 21 spikes were identified in $E_x$ and zero in $E_y$.

The only modifications made to the magnetic field measurements was the replacement of 36 data points in all components of the magnetic field on 2006-08-06 with linearly interpolated values. The values for all components were the same and much larger than the surrounding values.

@@

The GIC data contains non-physical spikes that are followed by a relaxation time of approximately 80 seconds. These spikes were removed by identifying times $t_s$ when the change in $|GIC|$ was greater than $0.04$ Amps and then data in a window $GIC(t_s-2)$ through $GIC(t_s+100)$ was replaced with a linear interpolation of the values at $GIC(t_s-3)$ and $GIC(t_s + 101)$. In the example interval shown in Figure~\ref{figure1}, 493 spikes were identified.

@@

To determine the coefficients $a_o$ and $b_o$ in the model (Model 1)

@@

A fourth model was also considered

@@

The frequency domain computations were performed on all possible unique 1-day window segments of available data, which, for Models 1, 2, and 4 resulted in 36 transfer functions as a function of the evaluation frequencies. The average transfer function magnitude was obtained by averaging the 36 transfer function magnitudes; the average transfer function phase was obtained using the average real and imaginary components, respectively, of the 36 transfer functions. The average impulse response functions were obtained by averaging the 36 impulse response functions.

There are many statistical methods that have been proposed to estimate transfer functions in the magnetotelluric literature that require additional steps, processing, and assumptions \citep{Egbert1986,Chave1987,Chave1989,Jones1989,Larsen1996,Egbert1997,Eisel2001,Chave2004,Chave2012,Chave2017} beyond the ones used that follow the basic method in \cite{Simpson2005}. The additional steps and processing were developed to address leakage, un-removed data spikes and shifts, high levels of noise contamination, time intervals in which the model assumptions are violated, bias, and the fact that errors in the fits of the transfer function at each evaluation frequency may not be Gaussian distributed. These works typically involve the estimation of a robust and low bias transfer function. In the results presented, we did not use a robust statistical method to calculate the transfer function or a bias correction algorithm - our primary interest is comparing the ability of different models to predict the data (in a mean-squared-error sense) rather than estimating low-bias model parameters. As discussed in ~\cite{Weigel2017}, a model developed using robust and bias correction methods will not necessarily be an optimal prediction model.

However, several alternative methods for computing the transfer functions were tested individually by modifying the basic method described above to determine if there was a change in the computed transfer functions and each model's prediction ability. When using these alternative methods, we find that the conclusions made regarding the differences each model's prediction ability do not change, and there are differences in the prediction ability that are within the calculated uncertainty, and the transfer function values do not change significantly at frequencies where the signal to noise ratio is above $\sim$5. The experiments in this regard are discussed in the following section.

@@

In order to compare models, we have computed the out-of-sample prediction error of each model on a given day by computing the parameters in Models 1, 2, and 4 using the remaining 35 days of data. This process was repeated to create 36 out-of-sample predictions for each model. The 95\% confidence limits in the average prediction efficiencys and correlation coefficients were calculated using a 1000 bootstrap samples of the 36 prediction errors.

\begin{table}
\caption{}
\centering
\begin{tabular}{l l}
\hline
Model & PE  \\
\hline
1\hspace{1em} $G_o(t) = a_oE_x(t) + b_oE_y(t)$ & 0.35 [0.25, 0.44] \\
2\hspace{1em} $G_E(t) = A(\omega)E_x(t) + B(\omega)E_y(t)$ & 0.61 [0.53, 0.64] \\
3\hspace{1em} $G_{E'}(t) = A(\omega)E'_x(t) + B(\omega)E'_y(t)$ & 0.78 [0.73, 0.83] \\
4\hspace{1em} $G_{B}(t) = Z_xB_x(t) + Z_yB_y(t)$ & 0.79 [0.75, 0.83] \\
\hline
\end{tabular}
\end{table}

@@

Based on the results in Table 1, we conclude that Model 1 is inferior to Model 2 in predicting GIC and the $PEs$ of Model 3 and 4 are statistically equivalent. Model 3, which uses magnetic field measurements as in input provides better estimates of GIC than Model 2. The last conclusion may be somewhat unexpected because Model 3 uses parameters from two different models. A proposed explanation is that the electric field measurements contain more noise than the magnetic field measurements.

The average parameters for Model 1 are $a_o =$ $71.9 \pm 27.8$ A/(V/km) and $b_o =$ $-70.0 \pm 65.8$ A/(V/km) and the distribution of the 36 parameters calculated using each 1-day interval is shown in Figure 2. The wide distribution and large error bars are suggestive that the model is a poor representation of the measurements. However, the model does have a good prediction performance - the PE is 0.35 and the data/model correlation coefficient of 0.68. In a similar way that $|\mathbf{B}(t)|$ or $d|\mathbf{B}(t)|/dt$ are often good predictors of $GIC(t)$ [REFS] despite the fact that physically $\mathbf{E}$ drives $GIC$ and $\mathbf{E}$ is related to $\mathbf{B}$ through a frequency dependent transfer function, Model 1 is a fair predictor of GIC, but is a poor approximation relative to Models 2 and 3.

The average parameters for Models 2 and 4 are shown in Figure~\ref{H}--\ref{Phi}. The error bars at each frequency or time is the standard error of the 36 values used to compute the average. Figure \ref{H} shows that the timescale of variations to impulse in $\mathbf{E}$ for Model 2 and $\mathbf{B}$ for Model 4 are similar.

@@

\item Averaging the 36 transfer functions using the Huber M-estimator with a psi-function \citep{Huber2011}.

\item Using weighted averages of the 36 transfer functions with the weighting at each frequency equal to the in-sample coherence at that frequency

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/plot_TF_aves_aobo_histograms-options-1.pdf}
\caption{Distribution of parameters in Model 1. Each set of parameters were computed 36 times using one day of 1-second cadence data. The dots indicate the average of each histogram: $a_o$ = 71.9 A/(V/km) and $b_o$ = -70.0 A/(V/km).}
\label{histogram}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/plot_TF_aves_SN-options-1.pdf}
\caption{}
\label{Phi}
\end{figure}

>>>>

Geomagnetically induced currents (GICs) are electric currents in man-made conducting systems due to electric fields induced near Earth's surface. These fields are a result of time variations in ionospheric currents on timescales on the order of hours \citep{Ohtani2000} and the movement of Earth's surface relative to slow-varying current systems in Earth's ionosphere that are near stationary relative to the Sun \citep{Stening2013}. Of particular interest are currents induced in electric power systems as they can lead to system degradation, disruption, and failure \citep{Albertson1993,NERC2012}. Accurate estimation of GICs given direct measurements of the electric field (or estimates based on the geomagnetic field) is important for power system design, retrospective analysis, and for the mitigation of the impacts of space weather on power systems [REFS].

A common assumption made in estimates of GIC using either measured or estimated values of the horizontal geoelectric field at Earth's surface, $\mathbf{E}$, is that the power system is resistive or quasi-DC so that the relationship $GIC(t) = a_oE_x(t) + b_oE_y(t)$ holds \citep{Albertson1981,Lehtinen1985}.  The coefficients $a_o$ and $b_o$ can then be calculated analytically using information about the connectivity of the power lines and values of the conductor and transformer resistances using DC circuit methods \citep[e.g.][]{Boteler2014a,Boteler2014b}. This assumption has been used or stated in many works, for example, \citet{Pulkkinen2007,Wik2008,Pulkkinen2010,Ngwira2011,Viljanen2012,Overbye2012}.

@@

The 1-second-cadence GIC data used in this paper span a total of 35 days across the 10 intervals: 2006-04-03 -- 2006-04-10, 2006-07-25 -- 2006-07-29, 2006-08-05 -- 2006-08-08, 2006-08-19 -- 2006-08-21, 2006-11-07 -- 2006-11-12, 2006-11-28 -- 2006-11-29, 2006-12-01, 2006-12-14 -- 2006-12-15, 2007-11-18 -- 2007-11-20, and 2007-11-22 -- 2007-11-22 that have been presented previously in \citet{Watari2009} and \cite{Watari2015}.

@@

Unphysical spikes in the electric field measurements assumed to be unphysical when the absolute value of a component $i$ of the electric field changed by more than 1 mV/km at time $t_s$, and the values $E_i(t_s-2)$ through $E_i(t_s+4)$ were replaced using linear interpolation of the values at $E_i(t_s-3)$ and $E_i(t_s + 5)$. In the sample shown in Figure~\ref{figure1}, 21 spikes were identified in $E_x$ and zero in $E_y$.

The only modifications made to the magnetic field measurements was the replacement of 36 data points in all components of the magnetic field on 2006-08-06 with linearly interpolated values. The values for $E_x$ and $E_y$ were the same and much larger than the surrounding values.

@@

The GIC data contains non-physical spikes that are followed by a relaxation time of approximately 80 seconds. These spikes were removed by identifying times $t_s$ when the change in $|GIC|$ was greater than $0.04$ Amps and then data in a window $GIC(t_s-2)$ through $GIC(t_s+100)$ was replaced using a linear interpolation of the values at $GIC(t_s-3)$ and $GIC(t_s + 101)$. In the example interval shown in Figure~\ref{figure1}, 493 spikes were removed.

@@

To determine the coefficients $a_o$ and $b_o$ in the frequency independent model, Model 1,

@@

A fourth model was also considered:

@@

Transfer function parameters were computed for each of the 35 unique 1-day window segments of available data. The average transfer function magnitude was obtained by averaging these 35 transfer function magnitudes; the average transfer function phase was obtained using the average real and imaginary components, respectively, of the 35 transfer functions. The average impulse response functions were obtained by averaging the 35 impulse response functions.

There are many statistical methods that have been proposed to estimate transfer functions in the magnetotelluric literature that use additional steps, processing, and assumptions \citep{Egbert1986,Chave1987,Chave1989,Jones1989,Larsen1996,Egbert1997,Eisel2001,Chave2004,Chave2012,Chave2017} beyond the ones used that follow the basic method in \cite{Simpson2005}. The additional steps and processing were developed to address leakage, un-removed data spikes and shifts, high levels of noise contamination, time intervals in which the model assumptions are violated, bias, and the fact that errors in the fits of the transfer function at each evaluation frequency may not be Gaussian distributed. These works typically involve the estimation of a robust and low bias transfer function. In the results presented, we did not use a robust statistical method to calculate the transfer function or a bias correction algorithm - our primary interest is comparing the ability of different models to predict the data (in a mean-squared-error sense) rather than estimating low-bias model parameters. As discussed in ~\cite{Weigel2017}, a model developed using robust and bias correction methods will not necessarily be an optimal prediction model.

However, several alternative methods for computing the transfer functions were tested individually by modifying the basic method described above to determine if there was a change in the computed transfer functions and each model's prediction ability. When using these alternative methods, we find that the conclusions made regarding the differences each model's prediction ability do not change, the differences in each model's prediction performance is within the calculated uncertainty of the used method, and the transfer function values are within the error bars of the used method when the signal to noise ratio is above $\sim$5. The experiments in this regard are discussed in the following section.

@@

The out-of-sample prediction metrics for each model on a given day was determined by computing the average model parameters using the remaining 34 days of data, generating a prediction time series on each of the 34 days using the average model parameters, and then computing the prediction metrics for each of the 34 days. This process was repeated to create a set of 35 out-of-sample prediction metrics for each model.

The performance of each model in predicting $GIC$ was assessed using three metrics: the prediction efficiency, PE, the mean-squared-error (MSE) ratio, and the signal-to-noise ratio.

@@

The average MSE ratio was computed by averaging the 35 MSE ratios and the average signal-to-noise ratio at a given frequency is the average of the 35 signal-to-noise ratios at that frequency.

The 95\% confidence interval, CI, on the  metrics was calculated using 1000 bootstrap samples of the 35 prediction metrics. The typical difference between the limits when the confidence intervals were computed assuming the values were gaussian-distributed was $\sim$2%.

\begin{table}
\caption{Model performance metrics for models $m=1-4$.}
\centering
\begin{tabular}{l c c c c}
\hline
$m$\hspace{1em} Model & PE & 95\% CI & MSE$_m$/MSE$_1$ & 95\% CI\\
\hline
1\hspace{1em} $G_o(t) = a_oE_x(t) + b_oE_y(t)$ & 0.37 & [0.31, 0.43] & 1.0 & \\
2\hspace{1em} $G_E(t) = A(\omega)E_x(t) + B(\omega)E_y(t)$ & 0.60 & [0.54, 0.65] & 1.8 & [1.6,2.0]\\
3\hspace{1em} $G_{E'}(t) = A(\omega)E'_x(t) + B(\omega)E'_y(t)$ & 0.78 & [0.76, 0.80] & 3.1 & [2.8,3.4]\\
4\hspace{1em} $G_{B}(t) = Z_xB_x(t) + Z_yB_y(t)$ & 0.83 & [0.80, 0.85] & 4.6 & [3.8,5.3]\\
\hline
\end{tabular}
\end{table}

The primary conclusion from the results shown in Table~1 is that the frequency dependent model, Model 2, is significantly better at predicting $GIC$ than the frequency independent model, Model 1.
Model 3, which uses magnetic field measurements as in input provides better estimates of $GIC$ than Model 2. The last conclusion may be somewhat unexpected because Model 3 uses parameters from two different models. A proposed explanation is that the electric field measurements contain more noise than the magnetic field measurements.

The average parameters for Model 1 are $a_o =$ $71.9 \pm 27.8$ A/(V/km) and $b_o =$ $-70.0 \pm 65.8$ A/(V/km) and the distribution of the 35 parameters calculated using each 1-day interval is shown in Figure~2. The wide distribution and large error bars are suggestive that the model is a poor representation of the measurements. However, the model does have a good prediction performance - the PE is 0.37 and the data/model correlation coefficient of 0.64. In a similar way that $|\mathbf{B}(t)|$ or $d|\mathbf{B}(t)|/dt$ are often good predictors of $GIC(t)$ [REFS] despite the fact that physically $\mathbf{E}$ drives $GIC$ and $\mathbf{E}$ is related to $\mathbf{B}$ through a frequency dependent transfer function, Model 1 is a fair predictor of $GIC$, but is a poor approximation relative to Models 2 and 3.

The average parameters for Models 2 and 4 are shown in Figure~\ref{H}--\ref{Phi}. The error bars at each frequency or time is the standard error of the 35 values used to compute the average. Figure \ref{H} shows that the timescale of variations to impulse in $\mathbf{E}$ for Model 2 and $\mathbf{B}$ for Model 4 are similar.

@@

\item Averaging the 35 transfer functions using the Huber M-estimator with a psi-function \citep{Huber2011}.

\item Using weighted averages of the 35 transfer functions with the weighting at each frequency equal to the in-sample coherence at that frequency

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/plot_TF_aves_SN-options-1.pdf}
\caption{Signal-to-Noise}
\label{SN}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/plot_TF_aves_aobo_histograms-options-1.pdf}
\caption{Distribution of parameters in Model 1. Each set of parameters were computed 35 times using one day of 1-second cadence data. The dots indicate the average of each histogram: $a_o$ = 71.9 A/(V/km) and $b_o$ = -70.0 A/(V/km).}
\label{histogram}
\end{figure}

@@
